SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-07 intro.tex
my pass

An ideal prefetching scheme would perfectly capture a program's memory access pattern, and then
predict and pre-load the needed data into the processor's caches in a timely manner. 
Memory access patterns may be simple, such as accessing every item in an array with a for-loop,
or very complex, such as chasing pointers through dynamically-allocated memory.
Practical prefetchers face challenges not only in predicting what data will be useful in the future,
but also in timing when data should be prefetched, deciding which level of the cache hierarchy
the data should be stored in, and deciding what data should be evicted from the caches to accommodate
the prefetched data.

All prefetchers are designed around a fundamental trade-off between two important metrics:
coverage and accuracy.  Prefetcher coverage refers to the number of cache hits that would have
been cache misses without the prefetcher.  For example, if an application experiences 1,000 cache misses
without a prefetcher, but only 200 cache misses with a prefetcher, then the prefetcher has 80\%
coverage for that application.  Prefetcher accuracy refers to the fraction of prefetched cache lines
that end up being used by the application.  So if a prefetcher prefetches 1,200 cache lines, 
but only 800 of them are used by the application, then that prefetcher's accuracy is 66.7\%.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{Motivation}
\caption{Variation of IPC, Total Prefetches and Useful Prefetches}
\label{Fig:Motivation}
\end{figure}

Coverage and accuracy are generally at odds with one another, and as one metric improves, the other
usually gets worse.  For example, when an application accesses a new region of memory for the first
time, a na\"ive prefetcher may predict that all data in that region will be used by the application.
This will clearly result in 100\% coverage for that region, but with possibly a very low accuracy.
In fact, so much bandwidth may be wasted prefetching unused data that performance can ultimately be
harmed by this strategy.  On the other extreme, another prefetcher may be overly
conservative and never prefetch anything, wasting no bandwidth, and achieving 0\% prefetch
coverage.

Figure~\ref{Fig:Motivation} illustrates the above scenario in a working manner.
Here we consider a state-of-the-art lookahead prefetcher -- SPP. We keep on 
re-tuning its thresholds, each time getting a variant with an increasing lookahead 
depth coverage. The figure depicts the behavior of the prefetcher variants running 
on the trace {\tt 603.bwaves\_s-1740B}. The IPC, total number of prefetches 
suggested by the prefetcher (TOTAL\_PF) and the actual useful predictions (GOOD\_PF),
all have been normalized to the least aggressive version of the prefetcher. 
As the look-ahead depth of the prefetch suggestions increase, so do useful prefetches. 
But that comes at a cost of total prefetches increasing at an even higher rate. This leads 
to cache pollution and bandwidth contention, as is clearly visible with the reduction in 
the IPC. Similar behavior was observed in 12 other SPEC CPU 2017 simpoints.

Therefore, a delicate balance between coverage and accuracy is required for a prefetcher to maximize
its performance impact.  Prefetchers are generally designed with internal mechanisms to
monitor their accuracy, with throttling thresholds that can be tuned for either coverage 
or accuracy.  The more irregular an application's memory access pattern is, the more difficult
it is to accurately predict every access, so a prefetcher will have to be tuned more toward
coverage (and away from accuracy) in order to gain any benefit.  This may be especially dangerous
to do in the context of a multi-core processor, where overly aggressive prefetching in one core can
waste shared resources, such as last-level cache (LLC) capacity, and off-chip bandwidth~\cite{Friendly}.

%Predicting future accesses requires a trade-off between coverage, the percent of predictions that 
%prevented a cache miss, and accuracy, the percent of predictions made that were correct. 
%%Gino: Just added a short explanation of coverage and accuracy tradeoff, not sure if necessary
%A prefetcher may have high coverage by reducing the number of misses by simply requesting a large 
%amount of data, however this does not imply accuracy. Even if the prefetcher brought in data accessed in the future,
%a high percentage of the data requested could never be accessed meaning the prefetcher had low accuracy, 
%resulting in wasted resources as well as cache pollution. Likewise, a prefetcher can have a 
%higher accuracy by being completely sure the data requested is used in the future, 
%but may not affect performance since it did not have the coverage needed to make an impact. %end addition.
%Most prefetchers maintain an internal confidence mechanism by keeping counters
%to track events such as cache hits, misses, references to an entry in the prefetcher's
%structure, etc. By comparing the confidence indicated by the counters against different 
%threshold values, the aggressiveness of the prefetcher can be adjusted, resulting in lower
%coverage and higher accuracy.

%For a prefetcher to capture highly irregular access patterns, it needs to be
%highly aggressive with high coverage. This causes the prefetcher to recommend many
%low accuracy prefetches that lead to the cache being polluted with useless data.  
%This effect is more pronounced in a multi-core scenario where the LLC is a
%shared resource~\cite{Friendly}.  Conversely, a conservative prefetcher would
%be highly accurate but might not make enough useful prefetches due to its lower coverage.

We propose Perceptron-based Prefetch Filtering (PPF) as an enhancement to
existing state-of-the-art prefetchers.  PPF works by observing the stream of
candidate prefetches generated by a prefetcher, and then rejects those that
are predicted by the online-trained neural model to be inaccurate.
The state-of-the-art prefetcher that we 
use to evaluate PPF in this paper is the Signature Path Prefetcher 
(SPP)~\cite{SPP}, however in principle, PPF could benefit any prefetcher.
In this design, PPF replaces SPP's existing confidence-based throttling mechanism,
which itself was a highly tuned feature of that prefetcher.  Because PPF is
so much more effective at rejecting inaccurate prefetches than SPP's baseline
mechanism, we are free to re-tune the rest of SPP's design around maximizing
coverage.  The result is an increase in both accuracy and coverage, and
a notable increase in performance.

@@

\item Implementing PPF filtering a state-of-the-art prefetchier, giving a
significant performance improvement compared to previous work.  PPF learns to
adapt itself to shared resource constraints, leading to further increased
performance in multi-core and bandwidth-constrained environments.

@@

In a single core configuration, PPF increases performance by 6.84\%
compared to the underlying prefetcher, SPP.  In a multi-core system running a
mixes of memory intensive SPEC CPU 2017 traces, PPF saw an improvement
of 11.9\% over SPP for a 4-core system, and 10.67\% for an 8-core system.

>>>>

An ideal prefetching scheme would perfectly capture a program's memory
access pattern, and then predict and pre-load the needed data into the
processor's caches in a timely manner.  Memory access patterns may be
simple, such as accessing every item in an array with a for-loop, or
very complex, such as chasing pointers through dynamically-allocated
memory.  % Practical prefetchers face challenges not only in predicting
% what data will be useful in the future, but also in timing when data
% should be prefetched, deciding which level of the cache hierarchy the
% data should be stored in, and deciding what data should be evicted
% from the caches to accommodate the prefetched data.
All prefetchers are designed around a fundamental trade-off between
two important metrics: coverage and accuracy. Prefetcher coverage
refers to the fraction of baseline cache misses that the prefetcher
pulls into the cache prior to their reference.  For example, if an
application experiences 1,000 cache misses without a prefetcher, while
800 of those cache misses become hits with a prefetcher, then the
prefetcher has 80\% coverage for that application.  Prefetcher
accuracy refers to the fraction of prefetched cache lines that end up
being used by the application. So if a prefetcher prefetches 1,200
cache lines, but only 800 of them are used by the application, then
that prefetcher's accuracy is 66.7\%.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{Motivation}
\caption{Variation of IPC, Total Prefetches and Useful Prefetches with
  lookahead depth.}
\label{Fig:Motivation}
\end{figure}

Coverage and accuracy are generally at odds with one another, and as
one metric improves, the other usually gets worse. For example, when
an application accesses a new region of memory for the first time, a
na\"ive prefetcher may predict that all data in that region will be
used by the application.  This will clearly result in 100\% coverage
for that region, but with possibly a very low accuracy.  In fact, so
much cache capacity and bandwidth may be wasted prefetching unused
data that performance can ultimately be harmed by this strategy. On
the other extreme, another prefetcher may be overly conservative and
never prefetch anything, wasting no capacity or bandwidth, and
achieving 0\% prefetch coverage.

Figure~\ref{Fig:Motivation} illustrates the above scenario.  Here we
consider a state-of-the-art. lookahead prefetcher -- SPP~\cite{SPP}.
Lookahead prefetchers such as SPP provide a mechanism to speculate an
arbitrary number of references ahead of the initial triggering miss.
In SPP, a throttling confidence threshold is then used to ensure that
the lookahead stops when confidence falls too low to ensure that
prefetches are accurate.  In the figure, we remove this threshold and
instead allow the prefetcher to lookahead an of arbitrary depth from 7
to 15.  The figure depicts the behavior of the {\tt 603.bwaves\_s}
SPEC CPU 2017 benchmark.  The IPC, total number of prefetches
suggested by the prefetcher (TOTAL\_PF) and the actual useful
predictions (GOOD\_PF), all have been normalized to the least
aggressive version of the prefetcher (depth 7).  As the look-ahead
depth increases, so do useful prefetches, and hence coverage.  This
coverage, however, comes at a cost of total prefetches increasing at
an even higher rate.  This leads to cache pollution and bandwidth
contention, as is clearly visible with the reduction in IPC.  Similar
behavior was observed in 12 other SPEC CPU 2017 benchmarks.

Therefore, a delicate balance between coverage and accuracy is
required for a prefetcher to maximize its performance impact.
Prefetchers are generally designed with internal mechanisms to monitor
their accuracy, and throttling mechanisms that can be tuned for either
coverage or accuracy.  The more irregular an application's memory
access pattern is, the more difficult it is to accurately predict
every access, so a prefetcher will have to be tuned more toward
coverage (and away from accuracy) in order to gain any benefit. This
may be especially dangerous to do in the context of a multi-core
processor, where overly aggressive prefetching in one core can waste
shared resources, such as last-level cache (LLC) capacity, and
off-chip bandwidth, impacting the performance of other
cores~\cite{Friendly}.

% Predicting future accesses requires a trade-off between coverage,
% the percent of predictions that prevented a cache miss, and
% accuracy, the percent of predictions made that were correct.
%% Gino: Just added a short explanation of coverage and accuracy
%% tradeoff, not sure if necessary
%A prefetcher may have high coverage by reducing the number of misses
% by simply requesting a large amount of data, however this does not
% imply accuracy. Even if the prefetcher brought in data accessed in
% the future, a high percentage of the data requested could never be
% accessed meaning the prefetcher had low accuracy, resulting in
% wasted resources as well as cache pollution. Likewise, a prefetcher
% can have a higher accuracy by being completely sure the data
% requested is used in the future, but may not affect performance
% since it did not have the coverage needed to make an
% impact. %end addition.
% Most prefetchers maintain an internal confidence mechanism by
% keeping counters to track events such as cache hits, misses,
% references to an entry in the prefetcher's structure, etc. By
% comparing the confidence indicated by the counters against different
% threshold values, the aggressiveness of the prefetcher can be
% adjusted, resulting in lower coverage and higher accuracy.

%For a prefetcher to capture highly irregular access patterns, it
% needs to be highly aggressive with high coverage. This causes the
% prefetcher to recommend many low accuracy prefetches that lead to
% the cache being polluted with useless data.  This effect is more
% pronounced in a multi-core scenario where the LLC is a shared
% resource~\cite{Friendly}. Conversely, a conservative prefetcher
% would be highly accurate but might not make enough useful prefetches
% due to its lower coverage.

Here, we propose Perceptron-based Prefetch Filtering (PPF) as an
enhancement to existing state-of-the-art prefetchers, allowing them to
speculate deeply to achieve high coverage while filtering out the
inaccurate prefetches this deep speculation implies.  PPF works by
observing the stream of candidate prefetches generated by a
prefetcher, and then rejects those that are predicted by the
online-trained neural model to be inaccurate.  The state-of-the-art
prefetcher that we use to evaluate PPF in this paper is the Signature
Path Prefetcher (SPP)~\cite{SPP}, however as we describe, PPF can be
designed to benefit any prefetcher.  In this design, PPF replaces
SPP's existing confidence-based throttling mechanism, which itself was
a highly tuned feature of that prefetcher.  Because PPF is so much
more effective at rejecting inaccurate prefetches than SPP's baseline
mechanism, we are free to re-tune the rest of SPP's design around
maximizing coverage. The result is an increase in both accuracy and
coverage, and a notable increase in performance.

@@

\item Implementing PPF filtering a state-of-the-art prefetcher, giving
  a significant performance improvement compared to previous work. PPF
  learns to adapt itself to shared resource constraints, leading to
  further increased performance in multi-core and
  bandwidth-constrained environments.

@@

In a single core configuration, PPF increases performance by 6.84\%
compared to the underlying prefetcher, SPP. In a multi-core system
running a mixes of memory intensive SPEC CPU 2017 traces, PPF saw an
improvement of 11.9\% over SPP for a 4-core system, and 10.67\% for an
8-core system.  \pg{update numbers before submit}
